---
title: "Modeling SPL"
author: "Becca Van Hoeck"
date: "March 21, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages and setwd

```{r}
setwd('C:/Users/Becca/Desktop/Soundscape_Analysis/Modeling')
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(mgcv))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(tidyr))
```

## Read in and prepare data

The response variable of interest is average sound pressure level (SPL_dB; loudness) measured in dB for each audio file. The dataframe also includes the linear values (spl_linear) so that I can summarize the data if needed without dealing with log rules. The data was subsampled to 2 minute audio files every 15 minutes, resulting in 96 files/day.The other variables included are:

hrmn_edt = Hour and minute in EDT of file (values range from 0:0.25:23.75)
year, month, day, hour, min, sec = individual columns of time of file in UTC
Site = Recording site: either 210RK, WESTR, CSPAR, or AEOLU 
Deploy = Possible values 1:5, 1 = Nov '15, 2 = Jan '16, 3 = Apr '16, 4 = June '16, 5 = Aug '16 

This only includes the low frequency data(fish sounds), I will likely do the same thing for the high frequency data (invertebrate sounds)

```{r echo=FALSE}
# The goal is to describe the daily pattern in SPL and evaluate how it varies across site and season


spl_data = read.csv('SPL_all_data.csv', header = TRUE)
spl_data = mutate(spl_data, Site = case_when(site == 1 ~ "210RK", site == 2 ~ "WESTR",
                                           site == 3 ~ "CSPAR", site == 4 ~ "AEOLU"),
                 Deploy = case_when(month == 11 ~ 1, month == 1 ~ 2, month == 4 ~ 3, 
                                    month == 6 ~ 4, month == 8 ~ 5), 
                 Reef_type = case_when(Site == "210RK" | Site == "WESTR" ~ "Natural",
                                       Site == "CSPAR" | Site == "AEOLU" ~ "Artificial"))
spl_data$id = paste(spl_data$Site,spl_data$Deploy, sep = "_")
spl_data$Deploy = as.factor(spl_data$Deploy)
spl_data$Site = as.factor(spl_data$Site)
spl_data = spl_data %>% group_by(id) %>% mutate(center_time = scale(matdays, center = TRUE))
spl_data$date = paste(spl_data$month, spl_data$day, sep = "_")
```

## Following GAM workflow from notes

```{r}
theme_set(theme_bw())
SPLdB_plot = ggplot(spl_data, aes(x = hrmn_edt, y = spl_low_db))+ geom_point()

# Plotting with different smoothers
SPLdB_plot+ geom_smooth(aes(color="Span=.667"),se=F)+ geom_smooth(aes(color="Span=.95"),se=F,span=.95)+labs(color="")+
  geom_smooth(aes(color="Span=.667, Robust Fit"),se=F,method.args=list(family="symmetric"))

# Simplest GAM
test_gam1 = gam(spl_low_db ~ s(hrmn_edt), data = spl_data)
SPLdB_plot+ geom_line(aes(y = predict(test_gam1)))

# Simple GAM, but plotting deployment and its interaction with site
test_gam2 = gam(spl_low_db ~ s(hrmn_edt) + Deploy*Site, data = spl_data)
ggplot(spl_data, aes(x = hrmn_edt, y = spl_low_db, color = Deploy, shape = Site))+ geom_point()+
  geom_line(aes(y = predict(test_gam2)))+ labs(color = "Deploy", shape = "Site")
summary(test_gam2)

```


## After meeting 04/19/2019

Using a GAM with a cyclic cubic basis to fit diurnal cycle.  This model accounts for variation among deployment in diurnal cycle and difference in site, deployment and day within deployment in the mean sound levels.  Note there is also a commented out model with a now correct smooth of `center_time` by each site by deployment interaction which would account for trends across the entire time. 


```{r}
## Successively reduce value of k
full_cc_gam = gam(spl_low_db ~ s(hrmn_edt, bs="cc", k=96,by = Deploy)+ factor(date)+
                    Site*Deploy, data=spl_data)
full_cc_gam1 = gam(spl_low_db ~ s(hrmn_edt, bs="cc", k=48,by = Deploy)+ factor(date)+
                    Site*Deploy, data=spl_data)
full_cc_gam2 = gam(spl_low_db ~ s(hrmn_edt, bs="cc", k=24,by = Deploy)+ factor(date)+
                    Site*Deploy, data=spl_data)
full_cc_gam3 = gam(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = Deploy)+ factor(date)+
                    id, data=spl_data)
gam.check(full_cc_gam3)

#Simplest model - R-sq only .11
cyclic_gamm = gamm(spl_low_db ~ s(hrmn_edt, bs = "cc", k=96, by = Deploy), data = spl_data)

## Each of these models successively reduces k value, the smallest k value provides a qualitatively similiar fit
## with reduced edf
cyclic_gamm1 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=96,by = Deploy)+ factor(date) + Site,
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))
cyclic_gamm2 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=48,by = Deploy)+ factor(date) + Site,
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))
cyclic_gamm3 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=24,by = Deploy)+ factor(date) + Site,
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))
cyclic_gamm4 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = Deploy)+ factor(date) + Site,
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))
# Singularity error
#cyclic_gamm5 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = Deploy)+ factor(date) + id,
#                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))

gam.check(cyclic_gamm4$gam)
par(mfrow = c(3,2))
plot(cyclic_gamm4$gam)

anova(cyclic_gamm1$lme,cyclic_gamm2$lme,cyclic_gamm3$lme,cyclic_gamm4$lme)
# The gam with the highest k has the highest edf, but the lowest AIC (cyclic_gamm1)
# GAM with lowest k has lowest edf, but highest AIC

cyclic_gamm5 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = interaction(Site,Deploy))+ factor(date), 
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))
cyclic_gamm6 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = interaction(Site,Deploy))+ factor(date) + Site,
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))
cyclic_gamm7 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = interaction(Site,Deploy))+ Site, 
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))

anova(cyclic_gamm5$lme,cyclic_gamm6$lme,cyclic_gamm7$lme)
# Based on anova results, cyclic_gamm6 has the lowest AIC and is selected for evaluating correlation structure 

anova(cyclic_gamm6$lme, cyclic_gamm4$lme)
summary(cyclic_gamm6$gam)
summary(cyclic_gamm4$gam)
# gamm6 has a lower AIC and a higher R-sq, continuing with that model


### Model errors with a singularity if Site*Deploy is included as a fixed effect
#cyclic_gamm8 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = interaction(Site,Deploy))+ factor(date) +
#                      Site*Deploy, data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))


#cont_cc_gam = gam(spl_low_db ~ s(hrmn_edt, bs="cc", k=96,by = Deploy)+ 
#                    s(center_time, by =interaction(Site,Deploy)) + Site*Deploy, data=spl_data)

```

## Investigating the GAMM

```{r}
summary(cyclic_gamm6$gam)
par(mfrow = c(5,4))
plot(cyclic_gamm6$gam)
gam.check(cyclic_gamm6$gam)

## Tried with k= 10 due to gam check results, but AIC increases - continuing with gamm6
cyclic_gammk10 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=8,by = interaction(Site,Deploy))+ factor(date) + Site,
                    data=spl_data, control = list(niterEM=0, optimMethod="L-BFGS-B"))
gam.check(cyclic_gammk10$gam)
plot(cyclic_gamm6$gam)

AIC(cyclic_gamm6,cyclic_gammk10)
anova(cyclic_gamm6$lme,cyclic_gammk10$lme)

# Exploring residuals and autocorrelation
filecount = as.vector(c(1:672,1:576,1:768,1:576,1:576,
                        1:672,1:768,1:768,1:672,1:672,
                        1:672,1:864,1:768,1:672,1:672,
                        1:768,1:672,1:672))

df_residuals = data.frame(filecount = filecount, ID = spl_data$id, residuals = residuals(cyclic_gamm6$lme)) %>%
  spread(ID,residuals,fill = NA)
par(mfrow=c(4,5))
acf(df_residuals[,2], lag.max = 50, na.action = na.omit) # 210 1
acf(df_residuals[,3], lag.max = 50, na.action = na.omit) # 210 2
acf(df_residuals[,4], lag.max = 50, na.action = na.omit) # 210 3 
acf(df_residuals[,5], lag.max = 50, na.action = na.omit) # 210 4
acf(df_residuals[,6], lag.max = 50, na.action = na.omit) # 210 5 
plot(x = 10, y = 20)
plot(x = 10, y = 20)
acf(df_residuals[,7], lag.max = 50, na.action = na.omit) # AEOLU 3
acf(df_residuals[,8], lag.max = 50, na.action = na.omit) # AEOLU 4
acf(df_residuals[,9], lag.max = 50, na.action = na.omit) # AEOLU 5
acf(df_residuals[,10], lag.max = 50, na.action = na.omit) # CSPAR 1
acf(df_residuals[,11], lag.max = 50, na.action = na.omit) # CSPAR 2
acf(df_residuals[,12], lag.max = 50, na.action = na.omit) # CSPAR 3
acf(df_residuals[,13], lag.max = 50, na.action = na.omit) # CSPAR 4
acf(df_residuals[,14], lag.max = 50, na.action = na.omit) # CSPAR 5
acf(df_residuals[,15], lag.max = 50, na.action = na.omit) # WESTR 1
acf(df_residuals[,16], lag.max = 50, na.action = na.omit) # WESTR 2 
acf(df_residuals[,17], lag.max = 50, na.action = na.omit) # WESTR 3
acf(df_residuals[,18], lag.max = 50, na.action = na.omit) # WESTR 4
acf(df_residuals[,19], lag.max = 50, na.action = na.omit) # WESTR 5
```

## Correlation structure

```{r}
#ctrl <- list(niterEM = 0, msVerbose = TRUE, optimMethod="L-BFGS-B")
ctrl = list(niterEM = 0, optimMethod="L-BFGS-B")

gamm6_AR = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = interaction(Site,Deploy))+ factor(date) + Site,
                    data=spl_data, correlation = corARMA(form = ~ 1|id, p = 1), control = ctrl)
anova(gamm6_AR$lme, cyclic_gamm6$lme)
plot(gamm6_AR$gam)
gam.check(gamm6_AR$gam)

df_residuals = data.frame(filecount = filecount, ID = spl_data$id, residuals = residuals(gamm6_AR$lme)) %>%
  spread(ID,residuals,fill = NA)
par(mfrow=c(4,5))
acf(df_residuals[,2], lag.max = 50, na.action = na.omit) # 210 1
acf(df_residuals[,3], lag.max = 50, na.action = na.omit) # 210 2
acf(df_residuals[,4], lag.max = 50, na.action = na.omit) # 210 3 
acf(df_residuals[,5], lag.max = 50, na.action = na.omit) # 210 4
acf(df_residuals[,6], lag.max = 50, na.action = na.omit) # 210 5 
plot(x = 10, y = 20)
plot(x = 10, y = 20)
acf(df_residuals[,7], lag.max = 50, na.action = na.omit) # AEOLU 3
acf(df_residuals[,8], lag.max = 50, na.action = na.omit) # AEOLU 4
acf(df_residuals[,9], lag.max = 50, na.action = na.omit) # AEOLU 5
acf(df_residuals[,10], lag.max = 50, na.action = na.omit) # CSPAR 1
acf(df_residuals[,11], lag.max = 50, na.action = na.omit) # CSPAR 2
acf(df_residuals[,12], lag.max = 50, na.action = na.omit) # CSPAR 3
acf(df_residuals[,13], lag.max = 50, na.action = na.omit) # CSPAR 4
acf(df_residuals[,14], lag.max = 50, na.action = na.omit) # CSPAR 5
acf(df_residuals[,15], lag.max = 50, na.action = na.omit) # WESTR 1
acf(df_residuals[,16], lag.max = 50, na.action = na.omit) # WESTR 2 
acf(df_residuals[,17], lag.max = 50, na.action = na.omit) # WESTR 3
acf(df_residuals[,18], lag.max = 50, na.action = na.omit) # WESTR 4
acf(df_residuals[,19], lag.max = 50, na.action = na.omit) # WESTR 5


gamm6_AR2 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=12,by = interaction(Site,Deploy))+ factor(date) + Site,
                    data=spl_data, correlation = corARMA(form = ~ 1|id, p = 2), control = ctrl)
anova(gamm6_AR2$lme, gamm6_AR$lme, cyclic_gamm6$lme)
plot(gamm6_AR2$gam)
gam.check(gamm6_AR2$gam)

gamm6_AR2_k24 = gamm(spl_low_db ~ s(hrmn_edt, bs="cc", k=24,by = interaction(Site,Deploy))+ factor(date) + Site,
                    data=spl_data, correlation = corARMA(form = ~ 1|id, p = 2), control = ctrl)
anova(gamm6_AR2$lme, gamm6_AR$lme, cyclic_gamm6$lme)

## gamm6_AR2 has the lowest AIC, but gam_check suggests that the k is too low. 
## Do I fit the AR2 structure to the model with a higher k? - this will take >4 hourss

df_residuals = data.frame(filecount = filecount, ID = spl_data$id, residuals = residuals(gamm6_AR2$lme)) %>%
  spread(ID,residuals,fill = NA)
par(mfrow=c(4,5))
acf(df_residuals[,2], lag.max = 50, na.action = na.omit) # 210 1
acf(df_residuals[,3], lag.max = 50, na.action = na.omit) # 210 2
acf(df_residuals[,4], lag.max = 50, na.action = na.omit) # 210 3 
acf(df_residuals[,5], lag.max = 50, na.action = na.omit) # 210 4
acf(df_residuals[,6], lag.max = 50, na.action = na.omit) # 210 5 
plot(x = 10, y = 20)
plot(x = 10, y = 20)
acf(df_residuals[,7], lag.max = 50, na.action = na.omit) # AEOLU 3
acf(df_residuals[,8], lag.max = 50, na.action = na.omit) # AEOLU 4
acf(df_residuals[,9], lag.max = 50, na.action = na.omit) # AEOLU 5
acf(df_residuals[,10], lag.max = 50, na.action = na.omit) # CSPAR 1
acf(df_residuals[,11], lag.max = 50, na.action = na.omit) # CSPAR 2
acf(df_residuals[,12], lag.max = 50, na.action = na.omit) # CSPAR 3
acf(df_residuals[,13], lag.max = 50, na.action = na.omit) # CSPAR 4
acf(df_residuals[,14], lag.max = 50, na.action = na.omit) # CSPAR 5
acf(df_residuals[,15], lag.max = 50, na.action = na.omit) # WESTR 1
acf(df_residuals[,16], lag.max = 50, na.action = na.omit) # WESTR 2 
acf(df_residuals[,17], lag.max = 50, na.action = na.omit) # WESTR 3
acf(df_residuals[,18], lag.max = 50, na.action = na.omit) # WESTR 4
acf(df_residuals[,19], lag.max = 50, na.action = na.omit) # WESTR 5


```

## Extracting Model terms 

```{r}
want <- seq(1, nrow(spl_data), length.out =3600)
pdat <- with(spl_data, data.frame(hrmn_edt = hrmn_edt[want], Site = Site[want],
                                  Deploy = Deploy[want], date = date[want]))
 
## predict trend contributions
p  <- predict(cyclic_gamm6$gam, newdata = pdat, type = "terms", se.fit = TRUE)
p1 <- predict(cyclic_gamm4$gam, newdata = pdat, type = "terms", se.fit = TRUE)

 
## combine with the predictions data, including fitted and SEs
pdat <- transform(pdat,
                 p  = p$fit[,2],  se  = p$se.fit[,2],
                 p1 = p1$fit[,2], se1 = p1$se.fit[,2])


op <- par(mar = c(5,4,2,2) + 0.1)
ylim <- with(pdat, range(p, p1))
ylim[1] <- floor(ylim[1])
ylim[2] <- ceiling(ylim[2])
ylab <- "Low Frequency SPL"
plot(spl_low_db- mean(spl_low_db) ~ hrmn_edt, data = spl_data, type = "n",
     ylab = ylab, ylim = ylim)
lines(p  ~ hrmn_edt, data = pdat, col = "black")
lines(p1 ~ hrmn_edt, data = pdat, col = "red")
par(op)

```


