---
title: "Modeling SPL"
author: "Becca Van Hoeck"
date: "March 21, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages and setwd

```{r}
setwd('C:/Users/Becca/Desktop/Soundscape_Analysis/Modeling')
library(ggplot2)
library(dplyr)
library(mgcv)
library(nlme)
library(tidyr)
```

## Read in and prepare data

The response variable of interest is average sound pressure level (SPL_dB; loudness) measured in dB for each audio file. The dataframe also includes the linear values (spl_linear) so that I can summarize the data if needed without dealing with log rules. The data was subsampled to 2 minute audio files every 15 minutes, resulting in 96 files/day.The other variables included are:

hrmn_edt = Hour and minute in EDT of file (values range from 0:0.25:23.75)
year, month, day, hour, min, sec = individual columns of time of file in UTC
Site = Recording site: either 210RK, WESTR, CSPAR, or AEOLU 
Deploy = Possible values 1:5, 1 = Nov '15, 2 = Jan '16, 3 = Apr '16, 4 = June '16, 5 = Aug '16 

This only includes the low frequency data(fish sounds), I will likely do the same thing for the high frequency data (invertebrate sounds)

```{r echo=FALSE}
spl_data = read.csv('SPL_all_data.csv', header = TRUE)
spl_data = mutate(spl_data, Site = case_when(site == 1 ~ "210RK", site == 2 ~ "WESTR",
                                           site == 3 ~ "CSPAR", site == 4 ~ "AEOLU"),
                 Deploy = case_when(month == 11 ~ 1, month == 1 ~ 2, month == 4 ~ 3, 
                                    month == 6 ~ 4, month == 8 ~ 5), 
                 Reef_type = case_when(Site == "210RK" | Site == "WESTR" ~ "Natural",
                                       Site == "CSPAR" | Site == "AEOLU" ~ "Artificial"))
spl_data$id = paste(spl_data$Site,spl_data$Deploy, sep = "_")
spl_data$Deploy = as.factor(spl_data$Deploy)
spl_data$Site = as.factor(spl_data$Site)
spl_data = spl_data %>% group_by(id) %>% mutate(center_time = scale(matdays, center = TRUE))

```

## Following GAM workflow from notes

```{r}
theme_set(theme_bw())
SPLdB_plot = ggplot(spl_data, aes(x = hrmn_edt, y = spl_low_db))+ geom_point()

# Plotting with different smoothers
SPLdB_plot+ geom_smooth(aes(color="Span=.667"),se=F)+ geom_smooth(aes(color="Span=.95"),se=F,span=.95)+labs(color="")+
  geom_smooth(aes(color="Span=.667, Robust Fit"),se=F,method.args=list(family="symmetric"))

# Simplest GAM
test_gam1 = gam(spl_low_db ~ s(hrmn_edt), data = spl_data)
SPLdB_plot+ geom_line(aes(y = predict(test_gam1)))

# Simple GAM, but plotting deployment and its interaction with site
test_gam2 = gam(spl_low_db ~ s(hrmn_edt) + Deploy*Site, data = spl_data)
ggplot(spl_data, aes(x = hrmn_edt, y = spl_low_db, color = Deploy, shape = Site))+ geom_point()+
  geom_line(aes(y = predict(test_gam2)))+ labs(color = "Deploy", shape = "Site")
summary(test_gam2)

```

## Models from meeting with James 

```{r}
# The goal is to describe the daily pattern in SPL and evaluate how it varies across site and season

## Fit gams that vary in their complexity of both smoothing effects and additive effects
#spl_gam1 <- gam(spl_low_db ~ s(hrmn_edt)+ Site*Deploy, data=spl_data)
#spl_gam2 <- gam(spl_low_db ~ s(hrmn_edt, by = Deploy) + Site*Deploy, data=spl_data)
full_addmod <- gam(spl_low_db ~ s(hrmn_edt, by = interaction(Site,Deploy)) + Site+Deploy, data=spl_data)
full_interactmod <- gam(spl_low_db ~ s(hrmn_edt, by = interaction(Site,Deploy)) + Site*Deploy, data=spl_data)
cont_interactmod <- gam(spl_low_db ~ s(matdays, by = interaction(Site,Deploy)) + Site*Deploy, data=spl_data)
#circ_interactmod <- gam(spl_low_db ~ s(sin((2*pi*matdays)/24) + cos((2*pi*matdays)/24), 
#                                       by = interaction(Site,Deploy)) + Site*Deploy, data=spl_data)


## Test for differences among nested models.
#anova(full_interactmod, full_addmod, test="Chisq")
#AIC(full_interactmod, full_addmod)
## ANOVA suggests that we keep the most complicated model
## The p-values are highly suspect given the likely temporal autocorrelation

summary(full_interactmod)
summary(cont_interactmod)
#summary(circ_interactmod)


## Using ggplot for quick gam visualization
ggplot(spl_data, aes(x=hrmn_edt, y=spl_low_db))+ 
  geom_point(alpha = .2)+
  geom_smooth(method="gam", formula= y~s(x))+ facet_grid(Site~Deploy)

ggplot(spl_data, aes(x = hrmn_edt, y = spl_low_db))+ geom_point(alpha = .2)+
  geom_line(aes(y = predict(full_interactmod)))+ facet_grid(Site~Deploy)
ggplot(spl_data, aes(x=matdays, y = spl_low_db))+ geom_point(alpha = .2)+
  geom_line(aes(y = predict(cont_interactmod)))+ facet_grid(Site~Deploy, scales = "free_x")

```

## Visualize and model Autocorrelation structure

```{r}
## Attempting acf on each site/deploy following convo with james
filecount = as.vector(c(1:672,1:576,1:768,1:576,1:672,
                        1:672,1:768,1:768,1:672,1:672,
                        1:672,1:864,1:768,1:672,1:672,
                        1:768,1:672,1:672))
df_residuals = data.frame(filecount = filecount, ID = spl_data$id, residuals = residuals(cont_interactmod)) %>%
  spread(ID,residuals,fill = NA)

#acf_obj = apply(df_residuals[,-1],2,acf, na.action = na.omit)
#df <- data.frame(matrix(unlist(acf_obj), nrow=length(acf_obj), byrow=T))
#test = do.call(rbind,acf_obj)

par(mfrow=c(4,5))
acf(df_residuals[,2], lag.max = 350, na.action = na.omit) # 210 1
acf(df_residuals[,3], lag.max = 350, na.action = na.omit) # 210 2
acf(df_residuals[,4], lag.max = 350, na.action = na.omit) # 210 3 
acf(df_residuals[,5], lag.max = 350, na.action = na.omit) # 210 4
acf(df_residuals[,6], lag.max = 350, na.action = na.omit) # 210 5 
plot(x = 10, y = 20)
plot(x = 10, y = 20)
acf(df_residuals[,7], lag.max = 350, na.action = na.omit) # AEOLU 3
acf(df_residuals[,8], lag.max = 350, na.action = na.omit) # AEOLU 4
acf(df_residuals[,9], lag.max = 350, na.action = na.omit) # AEOLU 5
acf(df_residuals[,10], lag.max = 350, na.action = na.omit) # CSPAR 1
acf(df_residuals[,11], lag.max = 350, na.action = na.omit) # CSPAR 2
acf(df_residuals[,12], lag.max = 350, na.action = na.omit) # CSPAR 3
acf(df_residuals[,13], lag.max = 350, na.action = na.omit) # CSPAR 4
acf(df_residuals[,14], lag.max = 350, na.action = na.omit) # CSPAR 5
acf(df_residuals[,15], lag.max = 350, na.action = na.omit) # WESTR 1
acf(df_residuals[,16], lag.max = 350, na.action = na.omit) # WESTR 2 
acf(df_residuals[,17], lag.max = 350, na.action = na.omit) # WESTR 3
acf(df_residuals[,18], lag.max = 350, na.action = na.omit) # WESTR 4
acf(df_residuals[,19], lag.max = 350, na.action = na.omit) # WESTR 5

```

## Exploring circular variables

```{r}
# GAM with circular time variable
circ_interactmod_centered = gam(spl_low_db ~ s(center_time, by = interaction(Site,Deploy))+ 
                            (sin(2*pi*center_time/.5) + cos(2*pi*center_time/.5))*id + Site*Deploy, data=spl_data)
summary(circ_interactmod_centered)
ggplot(spl_data, aes(x = center_time, y = spl_low_db))+ geom_point(alpha = .2)+
  geom_line(aes(y = predict(circ_interactmod_centered)))+ facet_grid(Site~Deploy)

# fit with lme too and look at residuals and subtract off daily cycle
circ_interactmod <- lm(spl_low_db ~ center_time + (sin(2*pi*center_time/24) + cos(2*pi*center_time/24))*id, 
                       data=spl_data)
summary(circ_interactmod)
dev.off()
plot(circ_interactmod)
ggplot(spl_data, aes(x = center_time, y = spl_low_db))+ geom_point()+
  geom_line(aes(y = predict(circ_interactmod)))+facet_grid(Site~Deploy)


circ_interactmod_coAR <- gls(spl_low_db ~ center_time+ (sin(2*pi*center_time/24) + cos(2*pi*center_time/24))*id,
                        correlation = corCAR1(form = ~center_time|id), data=spl_data)
summary(circ_interactmod_coAR)

# Evaluating remaining autocorrelation after accounting for daily pattern
df_residuals_circ = data.frame(filecount = filecount, ID = spl_data$id, 
                               residuals = residuals(circ_interactmod_coAR)) %>%
  spread(ID,residuals,fill = NA)
par(mfrow=c(4,5))
acf(df_residuals_circ[,2], lag.max = 350, na.action = na.omit) # 210 1
acf(df_residuals_circ[,3], lag.max = 350, na.action = na.omit) # 210 2
acf(df_residuals_circ[,4], lag.max = 350, na.action = na.omit) # 210 3 
acf(df_residuals_circ[,5], lag.max = 350, na.action = na.omit) # 210 4
acf(df_residuals_circ[,6], lag.max = 350, na.action = na.omit) # 210 5 
plot(x = 10, y = 20)
plot(x = 10, y = 20)
acf(df_residuals_circ[,7], lag.max = 350, na.action = na.omit) # AEOLU 3
acf(df_residuals_circ[,8], lag.max = 350, na.action = na.omit) # AEOLU 4
acf(df_residuals_circ[,9], lag.max = 350, na.action = na.omit) # AEOLU 5
acf(df_residuals_circ[,10], lag.max = 350, na.action = na.omit) # CSPAR 1
acf(df_residuals_circ[,11], lag.max = 350, na.action = na.omit) # CSPAR 2
acf(df_residuals_circ[,12], lag.max = 350, na.action = na.omit) # CSPAR 3
acf(df_residuals_circ[,13], lag.max = 350, na.action = na.omit) # CSPAR 4
acf(df_residuals_circ[,14], lag.max = 350, na.action = na.omit) # CSPAR 5
acf(df_residuals_circ[,15], lag.max = 350, na.action = na.omit) # WESTR 1
acf(df_residuals_circ[,16], lag.max = 350, na.action = na.omit) # WESTR 2 
acf(df_residuals_circ[,17], lag.max = 350, na.action = na.omit) # WESTR 3
acf(df_residuals_circ[,18], lag.max = 350, na.action = na.omit) # WESTR 4
acf(df_residuals_circ[,19], lag.max = 350, na.action = na.omit) # WESTR 5


```

## After meeting 04/19/2019

```{r}
mod1 = gam(spl_low_db ~ )
## getting code from james for gam with circular cubic spline

```


